Analyzing Fairness: Race, Recidivism, and the COMPAS Algorithm

### 2022-04-13 final_metrics_analysis.Rmd
get rid of blackboxauditing
updating metrics plot function

from aif360.sklearn.metrics import disparate_impact_ratio, average_odds_difference, statistical_parity_difference, equal_opportunity_difference
from aif360.sklearn.metrics import generalized_entropy_error, generalized_entropy_index, theil_index, consistency_score

@@

def plot_fair_metrics(fair_metrics_mitigated, model): 
  cols = ['statistical_parity_difference', 'equal_opportunity_difference', 'generalized_entropy']
  obj_fairness = [[0,0,1]]

@@

  # row for baseline model
  fair_metrics.loc['Baseline Model'] = [stat_par_diff, eq_opp_diff, gen_entr_error]

@@

  cols = fair_metrics.columns.values
  obj = fair_metrics.loc['objective']
  size_rect = [0.2,0.2,0.2,0.4,0.25]
  rect = [-0.1,-0.1,-0.1,0.8,0]
  bottom = [-1,-1,-1,0,0]
  top = [1,1,1,2,1]
  bound = [[-0.1,0.1],[-0.1,0.1],[-0.1,0.1],[0.8,1.2],[0,0.25]]

@@

## Individual Fairness Metrics

### Generalized entropy error

Generalized entropy index is proposed as a unified individual and group fairness measure. It measures the inequality in benefit allocation for individuals.
A value of 0 implies perfect fairness.
Fairness is indicated by lower scores, higher scores are problematic.

```{python group fairnesss generalized entropy error}
gen_entr_error = generalized_entropy_error(y_test, y_pred, alpha=1, pos_label = 0)

print(f'[Baseline] The generalized entropy error is {gen_entr_error: .2}')

#gen_entr_index = generalized_entropy_index(x_test, alpha=0, pos_label = 0)

#print(f'[Baseline] The generalized entropy index is {gen_entr_index: .2}')

#consistency_score = consistency_score(x_test, y_test, n_neighbors=5, pos_label = 0)

#print(f'[Baseline] The consistency score is {consistency_score: .2}')
```

```{python group fairness metrics plot}
plot_fair_metrics([0, 0, 0], '')
plt.show()
```

# Pre-Processing Approach
```{python pre-processing set up, include = FALSE}
lr = LogisticRegressionCV(solver='lbfgs')
rew = ReweighingMeta(estimator=lr, reweigher=Reweighing('race'))
rew.fit(X_train, y_train)
y_pred_REW = rew.predict(X_test)
```

@@

```{python pre-processing generalized entropy error}
gen_entr_error_RW = generalized_entropy_error(y_test, y_pred_REW, alpha=1, pos_label = 0)

print(f'[Reweighting] The generalized entropy error is {gen_entr_error_RW: .2}')
```

This is only a very slight improvement from 0.24 to 0.23.

```{python pre-processing fairness metrics plot}
plot_fair_metrics([stat_par_diff_RW, eq_opp_diff_RW, gen_entr_error_RW], 'Reweighting')
plt.show()
```

@@

diremover = DisparateImpactRemover(repair_level=1.0, sensitive_attribute="race")
#diremover.fit(X_train, y_train)
#y_pred_diremover = diremover.predict(X_test)

diremover_rp = diremover.fit_transform(X_test)
# AttributeError: 'DataFrame' object has no attribute 'features' (probem with .fit_transform function)
# .fit_transform() angry because of X_test...

>>>>

from aif360.sklearn.metrics import disparate_impact_ratio, average_odds_difference, statistical_parity_difference, equal_opportunity_difference

@@

def plot_fair_metrics(fair_metrics_mitigated, model): 
  cols = ['statistical_parity_difference','equal_opportunity_difference','average_odds_difference','disparate_impact_ratio']
  obj_fairness = [[0,0,0,1]]

@@

  # row for baseline model
  fair_metrics.loc['Baseline Model'] = [stat_par_diff, eq_opp_diff, avg_odds_diff, disp_impact_ratio]

@@

  cols = fair_metrics.columns.values
  obj = fair_metrics.loc['objective']
  size_rect = [0.2,0.2,0.2,0.4,0.25]
  rect = [-0.1,-0.1,-0.1,0.8,0]
  bottom = [-1,-1,-1,0,0]
  top = [1,1,1,2,1]
  bound = [[-0.1,0.1],[-0.1,0.1],[-0.1,0.1],[0.8,1.25]]

@@

```{python group fairness metrics plot}
plot_fair_metrics([0, 0, 0, 0], '')
plt.show()
```

# Pre-Processing Approaches

## Reweighing

```{python pre-processing set up, include = FALSE}
lr = LogisticRegressionCV(solver='lbfgs')
rew = ReweighingMeta(estimator=lr, reweigher=Reweighing('race'))
rew.fit(X_train, y_train)
y_pred_REW = rew.predict(X_test)
```

@@

```{python pre-processing average odds difference}
avg_odds_diff_RW = average_odds_difference(y_test, y_pred_REW, prot_attr='race', pos_label = 0)

print(f'[Reweighting] The average odds difference is {avg_odds_diff_RW: .2}')
```

This is also a vast improvement from -0.44 to 0.014. Since it is above 1 and below 0.1, it implies a slight benefit for Black defendants, but not an unfair benefit.

### Disparate Impact Ratio
```{python pre_processing disparate impact ratio}
disp_impact_ratio_RW = disparate_impact_ratio(y_test, y_pred_REW, prot_attr='race', priv_group = 1, pos_label = 0)

print(f'[Reweighting] The disparate impact ratio is {disp_impact_ratio_RW: .2}')
```

The reweighting approach also improves the disparate impact ratio, bringing the value almost to 1, which conveys absolute fairness.

```{python pre-processing fairness metrics plot}
plot_fair_metrics([stat_par_diff_RW, eq_opp_diff_RW, avg_odds_diff_RW, disp_impact_ratio_RW], 'Reweighting')
plt.show()
```

@@

# diremover = DisparateImpactRemover(repair_level=1.0, sensitive_attribute="race")
#diremover.fit(X_train, y_train)
#y_pred_diremover = diremover.predict(X_test)

# diremover_rp = diremover.fit_transform(X_test)
# AttributeError: 'DataFrame' object has no attribute 'features' (probem with .fit_transform function)
# .fit_transform() angry because of X_test...
